---
title: "Survival Analysis Sample Data Creation"
author: "Mick Cooney"
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
  pdf_document: default
---

<!--
(Title:) Survival Analysis Sample Data Creation

Author: Mick Cooney

Date: 2016

Abstract: This worksheet generate some very simple survival data for
the purposes of illustration of the technique.

Keywords: survival analysis, data generation

-->

```{r knit_opts, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

library(tidyverse)
library(data.table)
library(dtplyr)

library(feather)

library(survival)

library(GGally)
library(survminer)

library(cowplot)


options(width = 80L)

source("custom_functions.R")

set.seed(42)

theme_set(theme_gray())
```

# Create Example Survival Data

In this workbook we are going to generate a baseline hazard rate with
a time-varying effect and then add some proportional hazards in
addition to calculate some generated survival times. We will also add
some censorship.


```{r set_input_parameters, echo=TRUE}
N_dataset <- 10000

gender_dt <- data.table(gender      = c("M", "F")
                       ,gender_prop = c(0.65, 0.35)
                        )

health_dt <- data.table(health      = c("Unhealthy", "Average", "Healthy")
                       ,health_prop = c(        0.3,      0.55,      0.15)
                        )

early_rate <- 0.0180
late_rate  <- 0.0090


baseline_early <- rbeta(60, early_rate * 500, (1 - early_rate) * 500)
baseline_late  <- rbeta(60, late_rate  * 200, (1 - late_rate)  * 200)

baseline_hazard <- c(baseline_early, baseline_late)
baseline_hazard <- pmax(baseline_hazard, 0.0030)


gender_hazard <- 1.2
health_hazard <- c("Healthy" = 0.8, "Unhealthy" = 1.3)
age_hazard    <- function(age) 0.001 * age + 0.001 * age^2
```

Having determined the baseline hazard rate, we now create a plot of
the hazards. We will look at both the instantaneous hazards and the
cumulative survival rate.

```{r plot_input_rates, echo=TRUE}
ggplot() +
    geom_line(aes(x = seq_along(baseline_hazard), y = baseline_hazard)) +
    expand_limits(y = 0) +
    xlab("Month") +
    ylab("Instantaneous Chance of Failure")

ggplot() +
    geom_line(aes(x = c(0, seq_along(baseline_hazard))
                 ,y = cumprod(c(1, 1 - baseline_hazard)))) +
    expand_limits(y = 0) +
    xlab("Month") +
    ylab("Cumulative Survival Probability")
```

Now that we have set basic parameters to create the data, we now
generate some data and generate failure times. We also uniformly
generate observation times to ensure some of the values are censored.

```{r create_survival_inputs, echo=TRUE}
generate_fail_time <- function(probs) {
    fail_ids <- which(runif(length(probs) + 1, 0, 1) <= c(probs, 1))

    fail_ids[1]
}


# Generate covariate data and calculate the risk factors
sample_dt <- CJ(gender  = gender_dt$gender
               ,health = health_dt$health) %>%
    inner_join(gender_dt, by = 'gender') %>%
    inner_join(health_dt, by = 'health') %>%
    mutate(sample_prop = gender_prop * health_prop)

idx <- sample(1:nrow(sample_dt)
             ,size = N_dataset
             ,prob = sample_dt$sample_prop
             ,replace = TRUE)

data_dt <- sample_dt[idx] %>%
    mutate(age_at_start = round(rnorm(n(), 40, 10))
          ,rf_gender    = ifelse(gender == 'M', gender_hazard, 1)
          ,rf_health    = case_when(health == 'Average'   ~ 1
                                   ,health == 'Unhealthy' ~ 1.3
                                   ,health == 'Healthy'   ~ 0.8)
          ,rf_age       = age_hazard(age_at_start)
          ,risk_factor  = rf_gender * rf_health * rf_age
           )

glimpse(data_dt)
head(data_dt)
```

Now that we have our input data, we use this data to randomly generate
failure times, produce observation times based on a uniform draw and
then use this to construct failure times and censored observations.

```{r generate_survival_data, echo=TRUE}
data_dt$fail_time <- map_int(data_dt$risk_factor
                            , ~ generate_fail_time(. * baseline_hazard))

data_dt <- data_dt %>%
    mutate(obs_time = sample(seq_along(baseline_hazard), n(), replace = TRUE)
          ,time     = pmin(obs_time, fail_time)
          ,fail     = ifelse(fail_time <= obs_time, TRUE, FALSE)
           )

glimpse(data_dt)
head(data_dt)
```

With this data generated, we now look at some simple ratios for the
failure data we have.


```{r show_failure_data, echo=TRUE}
data_dt %>% group_by(fail) %>% summarise(count = n()) %>% print
```


# Create Kaplan-Meier Estimates

Now that we have a dataset to work with, we use it to perform some
simple descriptive exploration of the data.

We start with some Kaplan-Meier estimates of the cumulative survival.

```{r survival_km_nosplit, echo=TRUE}
nosplit_km <- survfit(Surv(time,fail) ~ 1, data = data_dt)
plot_1     <- ggsurvplot(nosplit_km, censor = FALSE, size = 0.5)

gender_km <- survfit(Surv(time,fail) ~ gender, data = data_dt)
plot_2    <- ggsurvplot(gender_km, censor = FALSE, size = 0.5)

health_km <- survfit(Surv(time,fail) ~ health, data = data_dt)
plot_3    <- ggsurvplot(health_km, censor = FALSE, size = 0.5)

both_km <- survfit(Surv(time,fail) ~ gender + health, data = data_dt)
plot_4  <- ggsurvplot(both_km, censor = FALSE, size = 0.5)


plot_grid(plot_1$plot, plot_2$plot, plot_3$plot, plot_4$plot, ncol = 2)
```



# Write to Disk

```{r write_sample_data, echo=TRUE}
write_feather(data_dt, path = 'data/sample_data.feather')

saveRDS(data_dt, 'km_dashboard/data_dt.rds', compress = 'xz')
```
